Best Practices for Multiple Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-05-24 paper/manuscript.tex
Improve phrasing

\section{\tocorange{Large-Scale Simulation Planning}}
\label{sec:large_scale_planning}
\tocsectioncomment{Possibly add discussion of methods which are not easily characterised as single/ multi e.g. lambda dynamics? Maybe a better distinction is calculations where we are interested in more than two end states.}
\tocsectioncomment{@OpenFE team to add discussion of how to run and adaptively update perturbation networks.}
\tocsectioncomment{Possibly add discussion of different protocols, especially active learning. Best practices may not be well established but many companies are running at scale.}
\tocsectioncomment{Discussion of merits of ABFE/ RBFE for virtual screening? See @Gary Tresadern comment from first meeting - lots of discussion around ABFE calculations for virtual screening.}
{\color{blue}
When tackling large datasets, methods that compute multiple ligands' binding free energies simultaneously offer an appealing alternative to pairwise calculations. These approaches can be categorized as "multiple end-state" free energy calculations (or "multistate" for short). They can in principle be more convenient than their pairwise counterparts as they sidestep ligand network construction entirely and can deliver all relative free energies in a single run. However, requiring many end states to be sampled at once complicates the energy landscape, forcing longer, inherently sequential simulations. Additionally they are often more difficult to set up: combining multiple ligand topologies raises extra challenges about avoiding bias from distance restraints or hybrid models while requiring sufficient phase-space overlap.

Enveloping distribution sampling \cite{christ2007enveloping} (EDS) works by constructing a single reference potential controlled by parameters that flatten inter-state barriers and balance end-state sampling. Multiple EDS potentials with different levels of barrier-smoothing can be combined in a replica-exchange setup in order to promote crossing the alchemical energy barriers that separate the end-states \cite{sidler2017efficient, champion2023accelerating}. This approach (RE-EDS) confers several practical advantages. Because the protein and solvent degrees of freedom are propagated only once per time step, the incremental cost of adding more ligands grows only linearly (for evaluating their potentials) rather than quadratically (in a fully connected perturbation graph). Furthermore, RE-EDS inherits intrinsic cycle closure from the enveloping reference state, eliminating the need to check or enforce thermodynamic cycles by hand.

Setting up RE-EDS simulations requires some thought to be put in system preparation. In the dual-topology setup all ligands must be restrained to avoid diffusing out of the binding site and ensuring phase-space overlap so that intermediate configurations smoothly connect ligand basins. However the number and definition of restraints should be chosen carefully as they can introduce bias in the simultations if not explicitely corrected. Multistate hybrid topologies can alleviate this issue as well as improve phase-space overlap \cite{champion2023accelerating} but their implementation in most MD engines usually requires custom routines, limiting their applicability.

RE-EDS workflows as implemented in the GROMOS \cite{champion2023accelerating} or OpenMM \cite{rieder2022leveraging} software packages include an automated calibration procedure in which the EDS potentials on the replica-exchange ladder are iteratively optimized by minimizing round-trip times and state occupancy. Nevertheless, heterogeneous datasets containing ligands that necessitate many heavy atom insertions or deletions can drastically reduce the efficiency of RE-EDS, and in these cases it is generally beneficial to split very large or poorly overlapping ligand sets into smaller groups or subsets to be executed with pairwise methods.
}

>>>>

\section{\tocorange{Large-Scale Simulation Planning}}
\label{sec:large_scale_planning}
\tocsectioncomment{Possibly add discussion of methods which are not easily characterised as single/ multi e.g. lambda dynamics? Maybe a better distinction is calculations where we are interested in more than two end states.}
\tocsectioncomment{@OpenFE team to add discussion of how to run and adaptively update perturbation networks.}
\tocsectioncomment{Possibly add discussion of different protocols, especially active learning. Best practices may not be well established but many companies are running at scale.}
\tocsectioncomment{Discussion of merits of ABFE/ RBFE for virtual screening? See @Gary Tresadern comment from first meeting - lots of discussion around ABFE calculations for virtual screening.}
{\color{blue}
When dealing with large datasets, methods that compute multiple ligands' binding free energies simultaneously offer a compelling alternative to pairwise calculations. These approaches, categorized as "multiple end state" or "multistate" free energy calculations, can be more convenient than standard methods as they sidestep ligand network construction entirely and can deliver all relative free energies in a single run. However, requiring many end states to be sampled at once complicates the energy landscape, forcing longer, inherently sequential simulations. Additionally, they are often more difficult to set up; combining multiple ligand topologies raises extra challenges about avoiding bias from distance restraints or hybrid models while requiring sufficient phase-space overlap.

A well-established method from this category is enveloping distribution sampling \cite{christ2007enveloping} (EDS), which constructs a single reference potential that samples the end state phase spaces of any number of states. Multiple EDS potentials with different levels of barrier-smoothing can be combined in a replica-exchange setup in order to promote crossing of alchemical energy barriers that separate the end-states \cite{sidler2017efficient, champion2023accelerating}. This approach (RE-EDS) confers several practical advantages. Because the protein and solvent degrees of freedom are propagated only once per time step, the computational cost of adding more ligands grows only linearly rather than quadratically (in a fully connected perturbation graph). Furthermore, RE-EDS inherits intrinsic cycle closure from the enveloping reference state, eliminating the need to check or enforce thermodynamic cycles by hand.

Setting up RE-EDS simulations requires some thought to be put in system preparation. In the dual-topology setup all ligands must be restrained to avoid diffusing out of the binding site and ensuring phase-space overlap so that intermediate configurations smoothly connect ligand energy wells. However the number and definition of restraints should be chosen carefully as they can introduce bias in the simultations if not explicitely corrected. Multistate hybrid topologies can alleviate this issue as well as improve phase-space overlap \cite{champion2023accelerating} but their implementation in most MD engines usually requires custom routines, limiting their applicability.

The RE-EDS workflow as implemented in the GROMOS \cite{champion2023accelerating} or OpenMM \cite{rieder2022leveraging} software packages includes an automated calibration procedure in which the EDS potentials on the replica-exchange ladder are iteratively optimized by minimizing round-trip times and balancing state occupancy. Nevertheless, heterogeneous datasets containing ligands that require insertion or deletion of many heavy atoms can drastically reduce the efficiency of the method, and in these cases it is generally beneficial to split very large or poorly overlapping ligand sets into smaller groups or subsets to be executed with pairwise calculations.
}
